Demers Cartogram with Rivers

### 2024-07-02 paper/main.tex
draft: revision

\keywords{Cartogram, Demers Cartogram}

@@

Among the four types of cartogram categorized in a survey by \citet{nusrat2016State} (See \Cref{sec:RelatedWork} for definitions of contiguous, non-contiguous, rectangular, and Dorling), a trade-off is made between types of accuracy (See \Cref{table:accuracy}).
For this project, we focus on non-contiguous cartograms like Demers cartograms because they facilitate statistical comparison between regions, they can make good use of screen space, and comparison of regions is useful when studying Electronic Health Records (EHR) data.
Demers cartograms offer the advantage in cases where the data is not directly correlated to region sizes.
In other words, Demers cartograms are useful when a data dimension is not describing the geography of the region it represents but is tied to something else, for example the health of its population.
Also, the comparison of magnitudes becomes an area estimation task, which is effective for a numeric data encoding \cite{munzner2014Visualization}.
See \citea{nickel2022Multicriteria} for a more complete description of the advantages that Demers cartograms offer.
Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy.
Standard Demers cartograms are composed of square nodes that representing geographic enumeration units.
As such, this can reduce their legibility.
We implement a new hybrid cartographic layout algorithm that combines rivers with the placement of the nodes representing geographic enumeration units, in this case, a Clinical Commissioning Group (CCG).
We hypothesize that introducing rivers improves the overall legibility of a cartogram.
By \textit{legibility} we mean readability and ability to interpret the cartogram.
To assess this hypothesis, we designed an experimental setup where participants engaged in correspondence and location tasks as part of a user study.
To reduce error and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout.
We then apply the algorithm to a real-world case study using EHR data to evaluate the result.
We present a user study that demonstrates its effectiveness.

@@

Our work extends the algorithm described by \citet{tong2018Cartograms} which incorporates a static topological feature into Demers cartograms.
Our work enhances that of \citet{tong2018Cartograms} in multiple ways.
First, we introduce multiple features (rivers) into the layout, as opposed to a single river.
Second, we make topological features dynamic and further improve legibility and geographical accuracy.
By the term \textit{dynamic}, we mean that the position of the rivers is updated as part of the layout algorithm.
In previous work, the river is static and serves merely as a boundary.
Third, we improve the algorithm to resolve stalemates.
Finally, the way we evaluate the cartograms is also different.
\citet{tong2018Cartograms} count river crossings to evaluate error (a statistical metric).
Here our focus is on readability, and thus we include a user study.

@@

{
    \begin{figure*}[tb!]
        \centering
        \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{flowchart.png}
        \caption{An overview of our hybrid layout algorithm incorporating rivers.
            See also \Cref{alg:UpdateLayout} in Section \ref{sec:main} for more detail.
            See \Cref{fig:flowchart-stalemate} for the logic of processing a stalemate.
            For illustration purposes, we show the rendered views alongside the logical views representing the actual computation.}
        \label{fig:flowchart}
    \end{figure*}
}

@@

Due to pandemic restrictions, we designed the user study to be carried out online.
Participants are expected to use their \new{retail} hardware for the tasks.
However, devices with small screens are not recommended because of the large size of the cartograms that are rendered.
The \new{within-subject} user study includes four parts:

@@

\textbf{P2:} The participants are given three practice tasks to familiarize themselves with the user study design.
A sample task is shown in \Cref{fig:task}, and accessible at \url{https://ghr.wangqiru.com/#/P1} and included in the online materials\footnotemark{}.
\new{The practice tasks are identital to the actual tasks, accompanied by instructions provided before the actual tasks commence.
The results from these practice tasks are not included in the final analysis.}
These tasks also serve as a quality test to see whether users are taking the study seriously.
A demonstration of three sample tasks is also included in the instructional video.

\textbf{P3:} The participants are asked to complete 16 location tasks that involve 4 targetCCGs.
\new{See the online materials for exact locations of target CCGs\footnotemark[\value{footnote}].}
Response and reaction time are recorded.
These 4 CCGs are selected to avoid extreme cases (thus bias the result), in terms of size, color, and location.

\textbf{P4:} The participants are asked to complete a questionnaire\footnotemark[\value{footnote}] that consists of \new{5-Point} Likert Scale questions and open-ended questions.

>>>>

\keywords{Cartogram, Demers Cartogram, EHR Visualization, Visual Analytics}

@@

Among the four types of cartogram categorized in a survey by \citet{nusrat2016State} (See \Cref{sec:RelatedWork} for definitions of contiguous, non-contiguous, rectangular, and Dorling), a trade-off is made between types of accuracy (See \Cref{table:accuracy}).
For this project, we focus on non-contiguous cartograms like Demers cartograms because they facilitate statistical comparison between regions, they can make good use of screen space, and comparison of regions is useful when studying Electronic Health Records (EHR) data.
Demers cartograms offer the advantage in cases where the data is not directly correlated to region sizes.
In other words, Demers cartograms are useful when a data dimension is not describing the geography of the region it represents but is tied to something else, for example the health of its population.
Also, the comparison of magnitudes becomes an area estimation task, which is effective for a numeric data encoding \cite{munzner2014Visualization}.
See \citea{nickel2022Multicriteria} for a more complete description of the advantages that Demers cartograms offer.
\new{One of the drawbacks of Demers' cartograms is that they may become more difficult to read when a region becomes displaced from its geospatial origin through the node layout process.
The layout of these more abstract shapes may simultaneously reduce the map's legibility and increase error.
See \citet{tong2018Cartograms} for a more detailed explanation.}
Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy.
Standard Demers cartograms are composed of square nodes that representing geographic enumeration units.
As such, this can reduce their legibility.
We implement a new hybrid cartographic layout algorithm that combines rivers with the placement of the nodes representing geographic enumeration units, in this case, a Clinical Commissioning Group (CCG).
We hypothesize that introducing rivers improves the overall legibility of a cartogram.
By \textit{legibility} we mean readability and ability to interpret the cartogram.
To assess this hypothesis, we designed an experimental setup where participants engaged in correspondence and location tasks as part of a user study.
To reduce error and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout.
We then apply the algorithm to a real-world case study using EHR data to evaluate the result.
We present a user study that demonstrates its effectiveness.

@@

Our work extends the algorithm described by \citet{tong2018Cartograms} which incorporates a static topological feature into Demers cartograms.
Our work enhances that of \citet{tong2018Cartograms} in multiple ways.
First, we introduce multiple features (rivers) into the layout, as opposed to a single river.
Second, we make topological features dynamic and further improve legibility and geographical accuracy.
By the term \textit{dynamic}, we mean that the position of the rivers is updated as part of the layout algorithm.
In previous work, the river is static and serves merely as a boundary.
\new{To illustrate this more clearly, we refer to the video demonstration published by \citet{tong2018Cartograms}. Because the behavior of the layout algorithm is dynamic, a video is more appropriate to convey the motivation of dynamic features of interest.
We can observe the cartographic layout algorithm at 1:30 and 2:40 of the video.
During the update process, we can clearly observe nodes cross the Thames River as the size expands.
As the size of all the nodes expand, a number of nodes are pushed south in order to make more use of screen space.
If we want to make the most efficient use of screen space, we need to translate the Thames River further south during the layout process in order to prevent nodes from crossing it.
This is one of the main motivations for introducing dynamic rivers. }
Third, we improve the algorithm to resolve stalemates.
Finally, the way we evaluate the cartograms is also different.
\citet{tong2018Cartograms} count river crossings to evaluate error (a statistical metric).
Here our focus is on readability, and thus we include a user study.

@@

{
    \begin{figure*}[tb!]
        \centering
        \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{flowchart.png}
        \caption{An overview of our hybrid layout algorithm incorporating rivers.
            See also \Cref{alg:UpdateLayout} in Section \ref{sec:main} for more detail.
            See \Cref{fig:flowchart-stalemate} for the logic of processing a stalemate.
            For illustration purposes, we show the rendered views alongside the logical views representing the actual computation. \new{All squares are the same size in this figure for illustration purposes.}}
        \label{fig:flowchart}
    \end{figure*}
}

@@

Due to pandemic restrictions, we designed the user study to be carried out online.
Participants are expected to use their retail hardware for the tasks.
However, devices with small screens are not recommended because of the large size of the cartograms that are rendered.
The within-subject user study includes four parts:

@@

\textbf{P2:} The participants are given three practice tasks to familiarize themselves with the user study design.
A sample task is shown in \Cref{fig:task}, and accessible at \url{https://ghr.wangqiru.com/#/P1} and included in the online materials\footnotemark{}.
The practice tasks are identital to the actual tasks, accompanied by instructions provided before the actual tasks commence.
The results from these practice tasks are not included in the final analysis.
These tasks also serve as a quality test to see whether users are taking the study seriously.
A demonstration of three sample tasks is also included in the instructional video.

\textbf{P3:} The participants are asked to complete 16 location tasks that involve 4 target CCGs.
See the online materials for exact locations of target CCGs.
Response and reaction time are recorded.
These 4 CCGs are selected to avoid extreme cases (thus bias the result), in terms of size, color, and location.

\textbf{P4:} The participants are asked to complete a questionnaire that consists of 5-Point Likert Scale questions and open-ended questions.
